The Need to Develop a Holistic Infectious Disease Surveillance System

### 2024-09-27 manuscript/manuscript.qmd
Update background and methods sections

At the heart of an outbreak detection system is a surveillance program built upon case detection, often utilizing individual diagnostic tests as necessary components of epidemiological investigations before the declaration of an outbreak [@murrayInfectiousDiseaseSurveillance2017; @zhouEliminationTropicalDisease2013; @pahoIntegratedApproachCommunicable2000; @worldhealthorganizationMeaslesOutbreakGuide2022; @craggOutbreakResponse2018].
When diseases have non-specific symptoms, without accurate measurement tools we cannot have confidence in our ability to correctly ascribe changes in symptoms within a population to a particular disease, and therefore detect outbreaks of specific pathogens.
As a result, it has been commonplace for surveillance systems to be developed around high-accuracy tests, such as Polymerase Chain Reaction (PCR) tests and immunoglobulin (Ig) tests, when financially and logistically feasible [@gastanaduyMeasles2019; @commissionerCoronavirusCOVID19Update2020; @grasslyComparisonMolecularTesting2020; @ezhilanSARSCoVMERSCoVSARSCoV22021; @worldhealthorganizationCholera2023; @essentialprogrammeonimmunizationepiimmunizationClinicalSpecimensLaboratory2018].
Depending on the disease in question, either sensitivity (the ability to correctly detect a true positive individual) or specificity (the ability to correctly discount a true negative) will be prioritized, as they often are at odds with each other [@westreichDiagnosticTestingScreening2019; @shrefflerDiagnosticTestingAccuracy2024; @parikhUnderstandingUsingSensitivity2008].
This balance is commonly defined within the Target Product Profile (TPP) of a test [@worldhealthorganizationTargetProductProfiles], which is a set of minimum characteristics that should be met for production and widespread use, helping to guide research and development.
For example, in the wake of the 2013 Ebola outbreak in Guinea a TPP was developed that listed the minimum acceptable sensitivity of 95% and specificity of 99% [@chuaCaseImprovedDiagnostic2015].
Recognizing that Ebola is not the major cause of fever and other non-specific symptoms in the region, it is arguably more important to prioritize the specificity of the disease, although the authors note that the severity of the infection requires a high level of sensitivity as the consequences of a missed case are dire at an individual and population level [@chuaCaseImprovedDiagnostic2015].

Much like the accuracy of an individual test, outbreak detection systems face the same issue regarding the prioritization of sensitive vs specific alerts [@germanSensitivityPredictiveValue2000; @worldhealthorganizationOperationalThresholds2014; @lewisTimelyDetectionMeningococcal2001].
For many disease systems, particularly in resource constrained environments where the burden of infectious diseases is highest [@gbd2019childandadolescentcommunicablediseasecollaboratorsUnfinishedAgendaCommunicable2023; @roserBurdenDisease2023], cases are counted and if they breach a threshold, be that weekly, monthly, or some combination of the two, an alert is triggered that may launch a further investigation and/or a response [@worldhealthorganizationMeaslesOutbreakGuide2022; @worldhealthorganizationOperationalThresholds2014].
In effect, this discretizes a distinctly continuous phenomenon into a binary measure, outbreak or no outbreak, for decision making purposes.
When considering reactive vaccination campaigns and non-pharmaceutical based interventions designed to reduce transmission and limit and suppress outbreaks through direct and indirect protection, early action has the potential to avert the most cases, and in some instances, can be the most cost effective as it provides the opportunity to incorporate later information within an “adaptive management” framework [@atkinsAnticipatingFutureLearning2020; @taoLogisticalConstraintsLead; @graisTimeEssenceExploring2008; @ferrariTimeStillEssence2014; @worldhealthorganizationConfirmingInvestigatingManaging2009; @minettiLessonsChallengesMeasles2013].
While this framing would point towards a sensitive surveillance system being optimal, with the benefits clearly defined in terms of morbidity and mortality avoided, each action comes with both direct and indirect financial and opportunity costs stemming from unnecessary activities that remove resources from future response activities.
Just as the balance of sensitivity and specificity of a test must be carefully evaluated, so must the balance at the outbreak level.

@@

By examining the combination of alert threshold and individual test characteristic in a modeling study that explicitly incorporates dynamical background noise, we aim to illustrate the need to develop a TPP for the whole detection system, not just one component.
To evaluate the alert system performance, we develop a set of outbreak definition criteria and surveillance metrics, drawing inspiration from acceptance sampling, ecological surveillance systems, and epidemiological surveillance systems guidelines and reviews [@yemshanovAcceptanceSamplingCosteffective2020; @christensenHerdlevelInterpretationTest2000; @muratoEvaluationSamplingMethods2020; @sternAutomatedOutbreakDetection1999; @calbaSurveillanceSystemsEvaluation2015; @guidelinesworkinggroupUpdatedGuidelinesEvaluating2001].
Using these metrics we overcome issues encountered by early warning systems that rely on dynamical values such as $R_{\text{effective}}$ in defining outbreaks [@jombartRealtimeMonitoringCOVID192021; @proverbioPerformanceEarlyWarning2022; @stolermanUsingDigitalTraces2023; @brettAnticipatingEpidemicTransitions2018; @salmonMonitoringCountTime2016], for example characterizing the end of an epidemic period is important in a timeseries where multiple outbreaks will occur.

@@

We constructed a stochastic compartmental non-age structured SEIR model of measles and simulated using a modified Tau-leaping algorithm with a time step of 1 day, utilizing binomial draws to ensure compartment sizes remained positive valued [@chatterjeeBinomialDistributionBased2005; @gillespieApproximateAcceleratedStochastic2001].
To stochastically reintroduce infections and cause recurring epidemics, commuter-style imports were added that are proportional to the size of the population and $R_{0}$ [@keelingModelingInfectiousDiseases2008], and the transmission parameter ($\beta_{t}$) is sinusoidal with a period of one year.
$R_{0}$ was set to 16, with a latent period of 10 days and infectious period of 8 days. The population was initialized with 500_000 individuals with Ghana-like birth and vaccination rates, and the final results were scaled up to the 2022 population size of Ghana (33 million). The full table of parameters can be found in @tbl-model-parameters. All simulations and analysis was completed in Julia version 1.10.3 [@bezansonJuliaFreshApproach2017], with all code stored at [https://github.com/arnold-c/OutbreakDetection](https://github.com/arnold-c/OutbreakDetection).

@@

#figure(
	table(
		columns: 3,
		..for ((label), (.., measles, nonmeasles)) in parameter_labels.zip(parameters) {
			(label, measles, nonmeasles)
		},
		[Importation rate], table.cell(colspan: 2, align: center, $(1.06*μ*R_0)/(√(N))$),
		[Population size (N)], table.cell(colspan: 2, align: center, "500_000, scaled to 33M"),
		[Initial proportion susceptible], table.cell(colspan: 2, align: center, "0.05"),
		[Initial proportion exposed], table.cell(colspan: 2, align: center, "0.0"),
		[Initial proportion infected], table.cell(colspan: 2, align: center, "0.0"),
		[Initial proportion recovered], table.cell(colspan: 2, align: center, "0.95"),
	)
)
```

@@

To examine the sensitivity of the detection system to background noise, we layered the measles incidence timeseries with one of four noise time series: Poisson-only noise, or dynamical noise with rubella-like parameters that could be in- or out-of-phase or with independent seasonality to the measles dynamics.
The total annual incidence of noise in each system was approximately 8 times the average annual measles incidence, with the dynamical noise systems also containing low levels of Poisson noise, reflecting Ghana-like dynamics.
Despite the average incidence of noise in each scenario being the same, the relative proportion of measles to noise varies differently throughout the timeseries, exacerbating the effects of imperfect diagnostic tests that will produce higher rates of false positives and negatives than ELISA-like diagnostics.
Imperfect tests are also more susceptible to incorrect results because of the varying prevalence throughout the timeseries, so the PPV of the system will not be static throughout the timeseries.
To account for the stochasticity of the model, we simulated 100 time series for each scenario, each lasting 100 years, before summarizing the distributions of outbreak detection methods.

@@

It is common to use expert review to define outbreaks when examining empirical data, but this is not feasible in a modeling study where tens of thousands of years are being simulated.
To account for this, many studies only simulate a single outbreak within a timeseries (repeating this short stochastic simulation multiple times to ensemble results), define an outbreak as a period where $R_{\text{effective}}$ > 1, or use a threshold of > 2 standard deviations (s.d.) over the mean seasonal incidence observed in empirical data [@sternAutomatedOutbreakDetection1999;; @jombartRealtimeMonitoringCOVID192021; @stolermanUsingDigitalTraces2023; @salmonMonitoringCountTime2016; @teklehaimanotAlertThresholdAlgorithms2004; @leclereAutomatedDetectionHospital2017].
Each method has its uses, but to evaluate the performance of an outbreak detection system in an endemic region where multiple sequential epidemics are expected it is important to clearly define the bounds of the outbreak, which can only be achieved by one of these methods (2 s.d. > mean).
This, however, assumes strong seasonal forcing and regular periodicity of incidence to produce a smooth enough baseline, which is not present as countries near measles elimination status [@grahamMeaslesCanonicalPath2019].
Here we define a true measles outbreak as a region of the timeseries that meets the following three criteria:

@@

Each of these is a necessary, but not sufficient, condition to the definition of an outbreak; all must be met.
For example, a region may produce 1000 total cases within the lower and upper bounds defined by the daily incidence being $\ge$ 5 cases, but if this period only lasts for 20 days it would not be considered an outbreak as it would not be possible to mount a response to.
The incidence of non-measles febrile rash (i.e., noise) does not affect the outbreak status of a region but does affect the alert status triggered by the testing protocol.

@@

The use of alert thresholds is common in locations burdened by measles, where access to high performance computing is limited making nowcasting-style approaches to outbreak detection impractical.
For this reason, we define an “alert” as any single day where the 7-day moving average of the daily incidence is greater than, or equal to, a threshold of X.
To examine the interaction of test and surveillance systems we vary the alert threshold, X, between 1 and 15 cases per day, before computing the evaluation metrics and identifying the optimal threshold for each individual test.
We also examined an alert method that required either the daily incidence or the 7-day moving average of daily incidence to exceed the alert threshold X (see supplemental results).

Each day, 60% of the measles and non-measles febrile rash cases visit the clinic for treatment, and Y% percentage of these clinic visits are tested, as all clinic visits are deemed to be suspected measles cases because they meet the clinical case definition.
This percentage of clinic visits, Y, that are tested is varied between 10% and 60%, in 10% increments, for all combinations of diagnostic test (apart from the clinical case definition) and alert threshold, defining the “testing scenario”.
While a reporting rate of 6-48% is high, we are only simulating a population of 500_000 individuals i.e., approximately 1.5% of Ghana’s population, which produces absolute incidence values that are consistent with observed reporting on the national scale. Each testing scenario uses one of the following tests:

- No test, where every suspected measles case is counted as positive for measles and used to trigger an outbreak alert, which is equivalent to the use of the clinical case definition. This is implemented as a diagnostic test with 100% sensitivity, 0% specificity, a 0-day lag in returning the result, and all individuals who visit the clinic are tested (not varied between 10-60%, like the other diagnostic tests described).
- An RDT equivalent with 85% sensitivity and specificity, and 0-day lag in result return. That is, 85% of true measles cases will be correctly labelled as positive, and 15% of non-measles febrile rash individuals that are tested will be incorrectly labelled as positive for measles. This acts as a lower bound of acceptability for a new measles RDT.
- An ELISA equivalent test with 100% sensitivity and specificity with a 0-day test result delay.
- An ELISA equivalent test with 100% sensitivity and specificity with a 7-day test result delay.
- An ELISA equivalent test with 100% sensitivity and specificity with a 14-day test result delay.

@@

For each of the previously defined tests, we calculate the number of test positive cases (that will include false positive and negative cases resulting from imperfect diagnostic tests), which is used to categorize the time series by alert status in conjunction with the alert threshold X.
The overlap with true outbreak status, or lack thereof, is used to classify the following metrics, and has been illustrated in @fig-outbreak-schematic:

- The percentage of alerts that are correct. This is analogous to the PPV of the alert system. Note that it is possible to trigger multiple alerts within a single outbreak, and each would be considered correct.
- The percentage of outbreaks that are detected. This is analogous to the sensitivity of the alert system.
- The detection delay. Note that if an alert precedes the start of the outbreak, as long as it successfully “captures” the outbreak within its bounds, it is considered to be correct, resulting in a negative detection delay i.e., an early warning triggered by false positives.
- The number of unavoidable and avoidable outbreak cases and deaths. Avoidable cases are defined as those that occur within an outbreak after a correct alert is first triggered i.e., cases that could theoretically be prevented with a perfectly effective and timely response. Unavoidable cases are the inverse: those that occur before a correct alert, or those that occur in an undetected outbreak. Using a fitted age distribution of measles cases in Ghana [@mintaProgressMeaslesElimination2023] and the age-specific case fatality ratio for Ghana in 2022 [@sbarraEstimatingNationallevelMeasles2023], we scaled the values for each one-year age cohort, per annum. In practice, not all cases deemed avoidable are (due to imperfect and delays in responses), but to minimize the sensitivity of the results to the response implementation and operational constraints we are counting them as such.

@@

The threshold that maximized outbreak detection accuracy given a testing scenario depends heavily on the noise structure.
For example, the optimal threshold for an RDT with 90% sensitivity/specificity and 40% testing of clinic visits was 4 cases per day with the dynamical in-phase noise, whereas it is 5 cases per day when Poisson-only noise was simulated (@tbl-optimal-thresholds).
This corresponds to an accuracy of 72% and 93%, respectively (@tbl-optimal-thresholds-accuracy).
Much higher accuracy can be achieved with imperfect tests under Poisson noise because large spikes of non-measles febrile rash do not occur that would lead to many false positives, triggering an erroneous alert.
With dynamical noise, this possibility can and does occur, and when the ratio of measles to non-measles cases is sufficiently low, drastically impacts the accuracy of an alert system.
This is why the values observed among ELISA tests are identical between the different noise structures: with a perfect test there are no false positive results.
For dynamical noise, RDTs never perform as well as an ELISA, even scenarios that include a 14-day lag between test procurement and result receipt.
However, with Poisson noise, once testing reaches 20% of clinic visits, equivalence can be observed at approximately 91% accuracy (@tbl-optimal-thresholds-accuracy).

@@

#figure(
	table(
		columns: 10,
		  fill: (x, y) => {
		    if y == 0 {gray}
		    if y == 1 {gray}
		  },
		[], table.cell(colspan: 2, align: center, "Test Characteristic"), table.cell(colspan: 7, align: center, "Testing Rate"),
		..optimal_thresholds.flatten()
	)
)
```

@@

The performance of an outbreak detection system is highly sensitive to the structure and level of background noise in the simulation.
Despite the mean daily noise incidence set to equivalent values between the dynamical and Poisson-only simulations, drastically different results are observed.
Under the assumption that non-measles febrile rash is relatively static in time (Poisson), RDTs can perform as well, if not better than ELISA tests at moderate to high testing rates, and at a fraction of the cost.
However, if it is expected that the noise is non-stationary, imperfect tests cannot overcome their accuracy limitations through higher testing rates, saturating at c. 74% accuracy, relative to ELISA’s 93%.

@@

For computational simplicity, this paper did not include demography in the model structure.
And while a simulation-based approach allows for complete determination of true infection status i.e., measle vs non-measles febrile rash cases, and therefore an accurate accounting of the outbreak and alert bounds, these simulations do not specifically represent any real-world setting.
The evaluation of empirical data does provide this opportunity, but at the cost of knowing the true infection status of individuals and confounding of multiple variables, limiting analysis to only those who are observed (i.e., not those in the community who do not visit a healthcare center), and removing the possibility to explore the sensitivity of the results to parameters of interest to a surveillance program e.g., testing rate, and the test itself.

@@

Given these limitations, the explicit values (i.e., optimal thresholds, accuracies etc.) should be interpreted with caution, and the exact results observed in the real-world will likely be highly dependent on unseen factors, such as the proportion of measles and non-measles sources of febrile rash that seek healthcare.
However, the general patterns should hold, and more importantly, the analysis framework provides a consistent and holistic approach to evaluating the tradeoff between individual level tests and the alert system enacted to detect outbreaks.

>>>>

At the heart of an outbreak detection system is a surveillance program built upon case detection, often utilizing individual diagnostic tests as necessary components of epidemiological investigations before the declaration of an outbreak [@murrayInfectiousDiseaseSurveillance2017; @zhouEliminationTropicalDisease2013; @pahoIntegratedApproachCommunicable2000; @worldhealthorganizationMeaslesOutbreakGuide2022; @craggOutbreakResponse2018].
For diseases with non-specific symptoms, accurate measurement tools are often necessary to confidently and correctly ascribe changes in symptom prevalence within a population to a particular disease, and therefore detect outbreaks of specific pathogens.
As a result, it has been commonplace for surveillance systems to be developed around high-accuracy tests, such as Polymerase Chain Reaction (PCR) tests and immunoglobulin (Ig) tests, when financially and logistically feasible [@gastanaduyMeasles2019; @commissionerCoronavirusCOVID19Update2020; @grasslyComparisonMolecularTesting2020; @ezhilanSARSCoVMERSCoVSARSCoV22021; @worldhealthorganizationCholera2023; @essentialprogrammeonimmunizationepiimmunizationClinicalSpecimensLaboratory2018].
Depending on the disease in question, either sensitivity (the ability to correctly detect a true positive individual) or specificity (the ability to correctly discount a true negative) will be prioritized, as they often are at odds with each other [@westreichDiagnosticTestingScreening2019; @shrefflerDiagnosticTestingAccuracy2024; @parikhUnderstandingUsingSensitivity2008].
This balance is commonly defined within the Target Product Profile (TPP) of a test [@worldhealthorganizationTargetProductProfiles], which is a set of minimum characteristics that should be met for production and widespread use, helping to guide research and development.
For example, in the wake of the 2013 Ebola outbreak in Guinea a TPP was developed that listed the minimum acceptable sensitivity of 95% and specificity of 99% [@chuaCaseImprovedDiagnostic2015].
Recognizing that Ebola is not the major cause of fever and other non-specific symptoms in the region, it is arguably more important to prioritize the specificity of the disease, although the authors note that the severity of the infection requires a high level of sensitivity as the consequences of a missed case are dire at an individual and population level [@chuaCaseImprovedDiagnostic2015].

Much like the accuracy of an individual test, outbreak detection systems face the same issue regarding the prioritization of sensitive vs specific alerts [@germanSensitivityPredictiveValue2000; @worldhealthorganizationOperationalThresholds2014; @lewisTimelyDetectionMeningococcal2001].
For many disease systems, particularly in resource constrained environments where the burden of infectious diseases is highest [@gbd2019childandadolescentcommunicablediseasecollaboratorsUnfinishedAgendaCommunicable2023; @roserBurdenDisease2023], cases are counted and if a pre-determined threshold is breached, be that weekly, monthly, or some combination of the two, an alert is triggered that may launch a further investigation and/or a response [@worldhealthorganizationMeaslesOutbreakGuide2022; @worldhealthorganizationOperationalThresholds2014].
In effect, this discretizes a distinctly continuous phenomenon into a binary measure, outbreak or no outbreak, for decision making purposes.
When considering reactive vaccination campaigns and non-pharmaceutical based interventions designed to reduce transmission and limit and suppress outbreaks through direct and indirect protection, early action has the potential to avert the most cases, and in some instances, can be the most cost effective as it provides the opportunity to incorporate later information within an “adaptive management” framework [@atkinsAnticipatingFutureLearning2020; @taoLogisticalConstraintsLead; @graisTimeEssenceExploring2008; @ferrariTimeStillEssence2014; @worldhealthorganizationConfirmingInvestigatingManaging2009; @minettiLessonsChallengesMeasles2013].
While this framing would point towards a sensitive surveillance system being optimal, with the benefits clearly defined in terms of morbidity and mortality avoided, each action comes with both direct and indirect financial and opportunity costs stemming from unnecessary activities that remove resources from future response capabilities.
Just as the balance of sensitivity and specificity of a test for an individual must be carefully evaluated, so must the balance at the outbreak level.

@@

By examining the combination of alert threshold and individual test characteristic in a modeling study that explicitly incorporates dynamical background noise, we aim to illustrate the need to develop a TPP for the whole detection system, not just one component.
To evaluate the alert system performance, we develop a set of outbreak definition criteria and surveillance metrics, drawing inspiration from acceptance sampling, ecological surveillance systems, and epidemiological surveillance systems guidelines and reviews [@yemshanovAcceptanceSamplingCosteffective2020; @christensenHerdlevelInterpretationTest2000; @muratoEvaluationSamplingMethods2020; @sternAutomatedOutbreakDetection1999; @calbaSurveillanceSystemsEvaluation2015; @guidelinesworkinggroupUpdatedGuidelinesEvaluating2001].
Using these metrics we overcome issues encountered by early warning systems that rely on dynamical values such as $R_{\text{effective}}$ in defining outbreaks [@jombartRealtimeMonitoringCOVID192021; @proverbioPerformanceEarlyWarning2022; @stolermanUsingDigitalTraces2023; @brettAnticipatingEpidemicTransitions2018; @salmonMonitoringCountTime2016], for example, characterizing the end of an epidemic period is important in a time series where multiple outbreaks will occur.

@@

We constructed a stochastic compartmental non-age structured SEIR model of measles and simulated using a modified Tau-leaping algorithm with a time step of 1 day, utilizing binomial draws to ensure compartment sizes remained positive valued [@chatterjeeBinomialDistributionBased2005; @gillespieApproximateAcceleratedStochastic2001].
To stochastically reintroduce infections and cause recurring epidemics, commuter-style imports were added that are proportional to the size of the population and $R_{0}$ [@keelingModelingInfectiousDiseases2008], and the transmission parameter ($\beta_{t}$) is sinusoidal with a period of one year and 20% seasonal amplitude.
$R_{0}$ was set to 16, with a latent period of 10 days and infectious period of 8 days. The population was initialized with 500,000 individuals with Ghana-like birth and vaccination rates, and the final results were scaled up to the approximate 2022 population size of Ghana (33 million) [@worldbankGhana]. The full table of parameters can be found in @tbl-model-parameters. All simulations and analysis was completed in Julia version 1.10.3 [@bezansonJuliaFreshApproach2017], with all code stored at [https://github.com/arnold-c/OutbreakDetection](https://github.com/arnold-c/OutbreakDetection).

@@

#table(
  columns: 3,
  ..for ((label), (.., measles, nonmeasles)) in parameter_labels.zip(parameters) {
	  (label, measles, nonmeasles)
  },
  [Importation rate], table.cell(colspan: 2, align: center, $(1.06*μ*R_0)/(√(N))$),
  [Population size (N)], table.cell(colspan: 2, align: center, "500,000, scaled to 33M"),
  [Initial proportion susceptible], table.cell(colspan: 2, align: center, "0.05"),
  [Initial proportion exposed], table.cell(colspan: 2, align: center, "0.0"),
  [Initial proportion infected], table.cell(colspan: 2, align: center, "0.0"),
  [Initial proportion recovered], table.cell(colspan: 2, align: center, "0.95"),
)
```

@@

To examine the sensitivity of the detection system to background noise, we layered the measles incidence time series with one of four noise time series structures: Poisson-only noise; or dynamical noise with rubella-like parameters that could be in- or out-of-phase, or with independent seasonality to the measles dynamics.
Only  dynamical in-phase noise and Poisson-only noise are presented in this manuscript.
For each noise structure, five different magnitudes of noise were simulated and evaluated: the total annual incidence of noise in each system was either the same magnitude as the average annual measles incidence, two times, four times, six times, or eight times the average annual measles incidence.
100 time series of 100 years were simulated for each scenario, before summarizing the distributions of outbreak detection methods.

@@

It is common to use expert review to define outbreaks when examining empirical data, but this is not feasible in a modeling study where tens of thousands of years are being simulated.
To account for this, many studies only simulate a single outbreak within a time series (repeating this short stochastic simulation multiple times to ensemble results), define an outbreak as a period where $R_{\text{effective}}$ > 1, or use a threshold of > 2 standard deviations (s.d.) over the mean seasonal incidence observed in empirical data (or from a 'burn-in' period of the simulation) [@sternAutomatedOutbreakDetection1999; @jombartRealtimeMonitoringCOVID192021; @stolermanUsingDigitalTraces2023; @salmonMonitoringCountTime2016; @teklehaimanotAlertThresholdAlgorithms2004; @leclereAutomatedDetectionHospital2017].
Each method has its uses, but to evaluate the performance of an outbreak detection system in an endemic region where multiple sequential epidemics are expected it is important to clearly define the bounds of the outbreak, which can only be achieved by 2 s.d. > mean ($R_\text{effective}$ will be less than 1 after an outbreak's peak, but still within what can be reasonably defined as the outbreak's bounds).
This, however, assumes strong seasonal forcing and regular periodicity of incidence to produce a smooth enough baseline, which is not present as countries near measles elimination status [@grahamMeaslesCanonicalPath2019].
Here we define a true measles outbreak as a region of the time series that meets the following three criteria:

@@

Each of these is a necessary, but not sufficient, condition to the definition of an outbreak; all must be met.
For example, a region may produce 1000 total cases within the lower and upper bounds defined by the daily incidence being $\ge$ 5 cases, but if this period only lasts for 20 days it would not be considered an outbreak as it would not be possible to mount a response to.
The incidence of non-measles febrile rash (i.e., noise) does not affect the outbreak status of a region but may affect the alert status triggered by the testing protocol.

@@

The use of alert thresholds is common in locations burdened by measles, where access to high performance computing is limited making nowcasting-style approaches to outbreak detection impractical.
For this reason, we define an “alert” as any single day where the 7-day moving average of the daily incidence is greater than, or equal to, a pre-specified alert threshold.
To examine the interaction of test and surveillance systems we vary the alert threshold, between 1 and 15 cases per day, before computing the evaluation metrics and identifying the optimal threshold for each combination of individual test, testing rate, noise structure, and noise magnitude.
We also examined an alert method that required either the daily incidence or the 7-day moving average of daily incidence to exceed the alert threshold (see supplemental results).

Each day, 60% of the measles and non-measles febrile rash cases visit the clinic for treatment, and a percentage of these clinic visits are tested, as all clinic visits are deemed to be suspected measles cases because they meet the clinical case definition.
This percentage of clinic visits that are tested is varied between 10% and 60%, in 10% increments, for all combinations of diagnostic test (except clinical case definition) and alert threshold, defining the “testing scenario”.
<!--While a reporting rate of 6-48% is high, we are only simulating a population of 500,000 individuals i.e., approximately 1.5% of Ghana’s population, which produces absolute incidence values that are consistent with observed reporting on the national scale.-->
Each testing scenario uses one of the following tests:

- No test, where every suspected measles case is counted as positive for measles and used to trigger an outbreak alert, which is equivalent to the use of the clinical case definition. This is implemented as a diagnostic test with 100% sensitivity, 0% specificity, a 0-day lag in returning the result, and all individuals who visit the clinic are tested (not varied between 10-60%, like the other diagnostic tests described).
- An RDT equivalent with 85% sensitivity and specificity, and 0-day lag in result return. That is, 85% of true measles cases will be correctly labelled as positive, and 15% of non-measles febrile rash individuals that are tested will be incorrectly labelled as positive for measles. This acts as a lower bound of acceptability for a new measles RDT.
- An RDT equivalent with 90% sensitivity and specificity, and 0-day lag in result return.
- An ELISA equivalent test with 100% sensitivity and specificity with a 0-day test result delay.
- An ELISA equivalent test with 100% sensitivity and specificity with a 7-day test result delay.
- An ELISA equivalent test with 100% sensitivity and specificity with a 14-day test result delay.

@@

For each of the previously defined tests, we calculate the number of test positive cases (that will include false positive and negative cases resulting from imperfect diagnostic tests), which is used to categorize the time series by alert status in conjunction with the alert threshold.
The overlap with true outbreak status, or lack thereof, is used to classify the following metrics, and has been illustrated in @fig-outbreak-schematic:

- The percentage of alerts that are 'correct'. This is analogous to the PPV of the alert system. Note that it is possible to trigger multiple alerts within a single outbreak, and each would be considered correct.
- The percentage of outbreaks that are detected. This is analogous to the sensitivity of the alert system.
- The detection delay. If an alert precedes the start of the outbreak, as long as it successfully “captures” the outbreak within its bounds, it is considered to be a 'correct' alert, resulting in a negative detection delay i.e., an early warning triggered by false positives.
- The number of unavoidable and avoidable outbreak cases and deaths. Avoidable cases are defined as those that occur within an outbreak after a correct alert is first triggered i.e., cases that could theoretically be prevented with a perfectly effective and timely response. Unavoidable cases are the inverse: those that occur before a correct alert, or those that occur in an undetected outbreak. Using a fitted age distribution of measles cases in Ghana [@mintaProgressMeaslesElimination2023] and the age-specific case fatality ratio for Ghana in 2022 [@sbarraEstimatingNationallevelMeasles2023], we scaled the values for each one-year age cohort, per annum. In practice, not all cases deemed avoidable are (due to imperfect and delays in responses), but to minimize the sensitivity of the results to the response implementation and operational constraints we are counting them as such.

@@

The threshold that maximized outbreak detection accuracy given a testing scenario depends heavily on the noise structure.
For example, when the ratio of average noise:measles incidence is 8, the optimal threshold for an RDT with 90% sensitivity/specificity and 40% testing of clinic visits was 4 cases per day with the dynamical in-phase noise, whereas it is 5 cases per day when Poisson-only noise was simulated (@tbl-optimal-thresholds).
This corresponds to an accuracy of 72% and 93%, respectively (@tbl-optimal-thresholds-accuracy).
Much higher accuracy can be achieved with imperfect tests under Poisson noise because large spikes of non-measles febrile rash do not occur that would lead to many false positives, triggering an erroneous alert.
With dynamical noise, this possibility can and does occur, and when the ratio of measles to non-measles cases is sufficiently low, drastically impacts the accuracy of an alert system.
This is why the values observed among ELISA tests are identical between the different noise structures: with a perfect test there are no false positive results.
For dynamical noise, RDTs never perform as well as an ELISA, even scenarios that include a 14-day lag between test procurement and result receipt.
However, with Poisson noise, once testing reaches 20% of clinic visits, equivalence can be observed at approximately 91% accuracy (@tbl-optimal-thresholds-accuracy).

@@

#table(
  columns: 10,
    fill: (x, y) => {
      if y == 0 {gray}
      if y == 1 {gray}
    },
  [], table.cell(colspan: 2, align: center, "Test Characteristic"), table.cell(colspan: 7, align: center, "Testing Rate"),
  ..optimal_thresholds.flatten()
)
```

@@

The performance of an outbreak detection system is highly sensitive to the structure and level of background noise in the simulation.
Despite the mean daily noise incidence set to equivalent values between the dynamical and Poisson-only simulations, drastically different results are observed.

> Despite the same average incidence of noise in each (comparable) scenario, the relative proportion of measles to noise varies differently throughout the time series, exacerbating the effects of imperfect diagnostic tests that will produce higher rates of false positives and negatives than ELISA-like diagnostics.
Imperfect tests are also more susceptible to incorrect results because of the varying prevalence throughout the time series, so the PPV of the system will not be static throughout the time series.

Under the assumption that non-measles febrile rash is relatively static in time (Poisson), RDTs can perform as well, if not better than ELISA tests at moderate to high testing rates, and at a fraction of the cost.
However, if it is expected that the noise is non-stationary, imperfect tests cannot overcome their accuracy limitations through higher testing rates, saturating at c. 74% accuracy, relative to ELISA’s 93%.

@@

For computational simplicity, this paper did not include demography in the model structure.
And while a simulation-based approach allows for complete determination of true infection status i.e., measles vs non-measles febrile rash cases, and therefore an accurate accounting of the outbreak and alert bounds, these simulations do not specifically represent any real-world setting.
The evaluation of empirical data does provide this opportunity, but at the cost of knowing the true infection status of individuals and confounding of multiple variables, limiting analysis to only those who are observed (i.e., not those in the community who do not visit a healthcare center), and removing the possibility to explore the sensitivity of the results to parameters of interest to a surveillance program e.g., testing rate, and the test itself.

@@

Given these limitations, the explicit values (i.e., optimal thresholds, accuracies etc.) should be interpreted with caution, and the exact results observed in the real-world will likely be highly dependent on unseen factors, such as the proportion of measles and non-measles sources of febrile rash that seek healthcare.
However, the general patterns should hold, and more importantly, the analysis framework provides a consistent and holistic approach to evaluating the trade-off between individual level tests and the alert system enacted to detect outbreaks.
